---
title: "Central Limit Theorem comparison for exponential distribution"
author: "yogarjun77"
date: "Thursday, May 21, 2015"
output: html_document
---


**Overview:**   An exponential distribution can be simulated in R using function rexp(n, lambda) where n is the rate parameter. Theoretically, both the mean and standard 
deviation of an exponential distribution is 1/lambda. A sample of 1000 simulations of the averages of 40 exponentials are generated and in this experiment and compared with the theoretical values and normal distribution.  


**Preparation: **  
a. Load libraries and determine parameters to be used
```{r preliminaries}
library(ggplot2)


lambda = 0.2  
n = 40  
sim = 1000 #total simulations  
set.seed(333)
```  
b. Simulation of sample distribution

```{r sample distribution}  
set.seed(333)  
exp2 = NULL   
for (i in 1 :sim){  
exp2 = c(exp2, mean(rexp(n,lambda)))  
}  

head(exp2)  
length(exp2)  
```
**1 - Sample mean vs Theoretical Mean**  
Mean of sample set of 1000 averages of 40 exponentials  
```{r sample mean}
  
        mean_exp2 <- round(mean(exp2),3)
        print(mean_exp2)  
```  
Theoretical mean(population mean)
```{r theoretical mean}  
        mean_pop <- round(1/lambda, 3)
        print(mean_pop)  
```  
The sample mean (5.045) is almost equal to the theoretical value(5.000), fulfilling the first condition for Central Limit Theorem E[X] = µ.

**2 - Sample Variance vs Theoretical Variance**  

Variance of sample set of 1000 averages of 40 exponentials:  
```{r sample var}  
        var_exp2 <- round(var(exp2),3)  
        print(var_exp2)  
```  

Theoretical variance (population variance) =  sigma^2^/n  
```{r theoretical var}  
        var_exp2 <- (round((((1/lambda)^2)/n),3))  
        print(var_exp2)  
```  

Again, the sample variance is almost equal to the theoretical value, fulfilling the second condition for CLT:     
                        Var[X] = sigma^2^/n  


**3 - Proving that the distribution is approximately normal**  

a)Comparing with a distribution of 1000 exponentials - would it be similar to a normal distribution?  

Create a sample distribution of 1000 exponentials

```{r 1000 exp}
set.seed(333)
exp_th <- rexp(1000, 0.2) 
hist(exp_th, col = "lightgreen", xlab = "Value", ylab = "frequency", main = "1000 exponential distribution")  
```  
  


Histogram for 1000 averages of 40 exponentials
```{r hist exp2}
hist(exp2, col = "lightblue", xlab = "value", ylab = "frequency", main = "Distribution of 1000 averages of 40 exponentials")
```

Mean & Variance of 1000 exponential 

```{r 1000 exp var and mean}
round(mean(exp_th),3)
round(var(exp_th)/1000, 3)
```

The histogram and variance values show that a distribution of 1000 exponentials are not similar to a normal distribution.  


b) Histogram of 1000 averages of 40 exponentials mapped onto a normal distribution

```{r histogram map normal}
hist(exp2, col= "lightblue", breaks = "sturges", freq = FALSE, main = "Sample distribution and density curve")
abline (v=mean_exp2, col = "darkblue", lwd = 2)
text(mean_exp2,0.4, mean_exp2)
lines(density(exp2), col="darkblue")
```  

The distribution exhibits a Gaussian shape almost similar to a normal distribution.  
  
    
c) Normal Q-Q plot  

```{r qqnorm}  
qqnorm(exp2, col = "darkblue")   
qqline(exp2, col = "black")  
```  
  
  
A Q-Q (quantile-quantile) plot is more effective to test if a sample is normally distributed. The resulting plots show that the normalized exp_2 (circles) closely fit the theoretical default normal (line).  


**Conclusion**  
The sample fulfilled 2 key conditions for the Central Limit Theorem Criteria  

a) E[X]=µ  
b) Var[X]=sigma^2^/n  

The histogram for 1000 averages of 40 exponentials(the sample) and its Q-Q plot positively indicate that the sample exhibits characteristics of a normal distribution according to the Central Limit Theorem.

Additionally just taking 1000 exponentials with the same random seed do not yield a similar distribution.

All these points show proof that the sample is highly similar to a normal distribution based on the Central Limit Theorem.